---
title: "Kaggle Quarto"
format: 
  html:
    df-print: paged
date: last-modified
editor: visual
toc: true
author:
  - name: Michael Oyewale
    orcid: 0000-0002-0167-9604
  - name: Johannes Schwenke
    orcid: 0009-0006-9795-6181
  - name: Thomas Stojanov
    orcid: 0000-0001-8172-5326
---

# TODO

-   Split cabin
-   Think about imputation / try first imputation
    -   imputation based on Title implemented, but for Miss do based on Parch –\> Johannes
    -   Double check functions –\> Johannes
-   Create histograms of variables and group by survived / didn't survive –\> Michael
-   Choice of pred model

# Log

## 2024-12-06

-   Data overview
-   Reformatting

## 2024-12-13

-   Reformatting and cleaning Quarto document & chunks
-   Added df-print to the Quarto file above, the html output reads much better

```{r packages}
#| warning: false
#| message: false
#| include: false

if(!require("tidyverse")){
  install.packages("tidyverse")
  library("tidyverse")                                                       
  # packages that you’re likely to use in everyday data analyses: ggplot2, dplyr, tidyr, readr, tibble, stringr, readr, forcats
}

if(!require("naniar")){
  install.packages("naniar")
  library("naniar")                                                          
  # assess and visualize missingness
}

if(!require("Hmisc")){
  install.packages("Hmisc")
  library("Hmisc")                                                          
  # assess and visualize missingness
}

if(!require("skimr")){
  install.packages("skimr")
  library("skimr")                                                          
}

if(!require("tidymodels")){
  install.packages("tidymodels")
  library("tidymodels")                                                          
  # modelling
}
```

```{r custom-functions}
source("functions/mean_imput_missing_age.R")
source("functions/clean_data.R")
```

# Introduction and setup {#sec-introduction}

In the present Quarto document, we report our commonly developed solution to obtain predictions as part of a Kaggle competition (<https://www.kaggle.com/competitions/titanic/code>).

## Expected output {#sec-expected-output}

In short, the aim of the competition is to predict, as accurately as possible, the survival status of passengers of Titanic using individual data (*i.e.*, name, age, sex, socio-economic status, etc...).

For this, Kaggle competitors have to upload a dataset composed of two variables: - `PassengerId`: describes the passenger Id - `Survived`: describes the observed survival status of the passenger. This corresponds to the outcome to be predicted using the developed prediction model.

We load the example provided by the Kaggle competition authors using `read.csv()`:

```{r load submission}
# Example of what a submission file should look like
submission_example <- read.csv("data/gender_submission.csv")
head(submission_example, n = 10)
```

## Available datasets

Two datasets are made readily available by the Kaggle competition authors.

### Test set

The set of `PassengerId` expected in the output (see [-@sec-expected-output]) actually corresponds to the set of `PassengerId` of the so-called `test` dataset.

The `test` dataset is usually the dataset used to evaluate the validity of a prediction model. We load it using `read.csv()`:

```{r load test}
test <- read.csv("data/test.csv") # test data
test |> head(n = 10)
```

### Train set

When we develop a prediction model, we usually train the prediction model using the so-called `train` data set. We load this data using again `read.csv()`:

```{r load train}
train <- read.csv("data/train.csv") # train data
train |> head(n = 10)
```

**Remarks**: It looks like the `train` and `test` datasets are composed of the same number of variables. It is not necessarily always the case. Indeed, while variables could be available in a given data set, they are not always available in another data set. Therefore, each variable used in a prediction model should be carefully inspected both for availability and completeness in the two `training` and the `test` sets. We provide a more detailed inspection in the following section.

# Data inspection (pre cleaning)

We already had a look a the training data above using `head()`. Let's inspect some more.

In the following chunk, using `str()`, we have a look at the dataset, and the variables composing it, and more specifically we want to check their format:

```{r inspect2 train}
str(train)
```

**Remarks**:

-   `Survived`, `Pclass`, `Sex` and `Embarked` are currently stored as integers or characters, while they should be factor variables. We address this issue in @sec-convert.

-   `Cabin` and `Embarked` variables contain empty character strings. We address this issue in @sec-replace.

-   `Ticket` variable seem to be messy but might contain important information. We address this issue in @sec-inspection-Ticket.

We provide an initial variables distribution using `summary()`:

```{r inspect3 train}
summary(train)
```

We provide an initial description of the number of missing values., with `gg_miss_var()`:

```{r}
gg_miss_var(train, show_pct = TRUE) #very likely to be incorrect, that only Age has missing values
```

We provide first an inspection of `Ticket`:

```{r inspection-ticket-head} data1 |> pull(Ticket) |> head(n = 20)}
```

**Remarks:** While it seems the data are not structured, it looks like there are two parts in the `Ticket` variable: a prefix and a ticket number. We will separate them in during data cleaning.

# Data management

## Data cleaning

```{r}
train_clean <- train |> clean_data()
```

`clean_data` does the following:

-   Convert `Survived`, `Pclass`, `Sex`, `Embarked` to factor

-   Replace empty string in `Cabin` and `Embarked` with NA

-   Extract ticket prefix and ticket number from `Ticket` and store as `ticket_nr` and `ticket_pre`

We provide again a description of the number of missing values., with `gg_miss_var()`:

```{r}
gg_miss_var(train_clean, show_pct = TRUE)
```

**Remarks:**

-   `Cabin`: has a high proportion of missing
-   `Age`: has a high number of missing.

# Inspection after cleaning

## Inspection of `Ticket` variable {#sec-inspection-Ticket}

#### Relation between prefix and Cabin?

Is there any association between ticket pre and the type of Cabin?

```{r}
train_clean |> 
  dplyr::select(PassengerId, Cabin, ticket_pre)|> 
  group_by(ticket_pre, Cabin) |> 
  arrange(Cabin) |> 
  count() |> 
  dplyr::mutate(
    prop = n/sum(n)
  )
```

```{r}
train_clean |> 
  group_by(Pclass, ticket_pre) |> 
  dplyr::count()
```

```{r inspect11}
train_clean |> pull(ticket_pre) |> head(n = 20) 

#See below for possible categorization and freq count
```

## Inspection of `Cabin` variable {#sec-inspection-Cabin}

```{r}
data1 <- train_clean |>
  mutate(Cabin_sector = ifelse(!is.na(Cabin), str_extract(Cabin, "[A-GT](?=[^A-GT]*$)"), NA)) # Extracts last occurrence of letter in Cabin string
# --> Sector is looked at below together with Pclass
```

```{r}
data1 <- data1  |> 
  mutate(CabinNumber = 
           ifelse(!is.na(Cabin), #only process non-NA values
                  sapply(str_extract_all(Cabin, "\\d+"), function(x) { 
                    # Iterates through the list of extracted numbers (x)
                    # "B23 B25 B27" → c("23", "25", "27")  
                    if (length(x) > 0) {
                      max(as.numeric(x)) # Calculate max() and convert to numeric format
                    } else {NA
                        }
                    }
                    ),
                  NA
                  )
         ) #Extracted highest number from Cabin; NA if no number
```

```{r}
data1$CabinNumber |> head(n = 10)
```

## Inspection of `Name` variable {#sec-inspection-Name}

```{r inspect13}
data1 |> arrange(Name) |> head(Name, n = 10) #There are some additional names listed for some patients - Why is this? 
# (...) These names don't refer to other patients on the ship - don't know what these are
# \"...\" - most probably a nickname
# (\"...\") - don't know



# What to do with sibsp / parch?
# Should we have variable 'Family' (0/1) instead of sibsp / parch? 
# Or is one better for survival than the other? 
# E.g., will I be safer if I have siblings? Or will I be at a higher risk if I have a child to care after?

# sibsp: The data1set defines family relations in this way...
  # Sibling = brother, sister, stepbrother, stepsister
  # Spouse = husband, wife (mistresses and fiancés were ignored)
# parch: The data1set defines family relations in this way...
  # Parent = mother, father
  # Child = daughter, son, stepdaughter, stepson
  # Some children travelled only with a nanny, therefore parch=0 for them.
```

## Data visualization

## Inspection of age

```{r}
data1 |> # need to address the missing values
  filter(
    !is.na(Survived)
  ) |> 
  ggplot(aes(x = Age)) +
  geom_histogram(binwidth = 2,
                 alpha = 0.5,
                 position = "identity")
```

### Histograms

```{r}
data1 |> 
  filter(!is.na(Survived) & !is.na(Age)) |> 
  ggplot(aes(x = Age, fill = Survived, color = Survived)) +
    geom_histogram(binwidth = 2,
                   alpha = 0.5,
                   position = "identity")
```

```{r}
# Age in years
# Age is fractional if less than 1. 
# If the age is estimated, is it in the form of xx.5
# Move age_estimated to new column ?
hist(data1$Age)

# Passenger fare
hist(data1$Fare, breaks = 50)

# Nr of siblings / spouses aboard the Titanic
# The data1set defines family relations in this way...
# Sibling = brother, sister, stepbrother, stepsister
# Spouse = husband, wife (mistresses and fiancés were ignored)
hist(data1$SibSp)

# Nr of parents / children aboard the Titanic
# The data1set defines family relations in this way...
# Parent = mother, father
# Child = daughter, son, stepdaughter, stepson
# Some children travelled only with a nanny, therefore parch=0 for them.
hist(data1$Parch)
```

### Frequency tables

```{r}
# For data1$Sex
table(data1$Sex, useNA = "ifany")

# For data1$Pclass - A proxy for socio-economic status (SES)
table(data1$Pclass, useNA = "ifany")

# For data1$Embarked
table(data1$Embarked, useNA = "ifany")

#for outcome (death)
table(data1$Survived, useNA = "ifany")

#for outcome (cabin)
table(data1$Cabin, useNA = "ifany")

#for newly created column 'Prefixes'
table(data1$ticket_pre, useNA = "ifany")
# Could theoretically clean this column.. 
# A./5., A.5., A/5, A/5. --> A/5
# SOTON/O.Q., SOTON/O2, SOTON/OQ, STON/O, STON/O2. --> SOTON 
# and so on... and see whether it correlates with anything... 
# Care: 665 missing values
```

```{r}
data1 |> 
  dplyr::select(
    PassengerId,
    Pclass,
    Cabin,
    Cabin_sector,
    Ticket,
    ticket_nr,
    ticket_pre,
  ) |> 
  dplyr::group_by(
    Pclass
  ) |> 
  count(
    Cabin_sector
  ) |> 
  mutate(
    prop = n/sum(n)
  )
```

```{r}
data1 |> count(Pclass, Cabin_sector)
# Pclass doesn't necessarily separate Cabin_sector

data1 |> count(Pclass, ticket_pre)
# No real relation here either - check again
```

```{r}
data1 |> 
  group_by(Pclass) |> 
  count(ticket_pre) |> 
  arrange(desc(n))
```

```{r}
data1 |>
  group_by(Cabin_sector) |> 
  count(ticket_pre)
```

# Missing values

```{r}
gg_miss_var(data1, show_pct = TRUE)


naplot(naclus(data1))
```

# Imputation

The in the `Name` variable is always followed by a `.` –\> we can easily extract using regex and create an overview of the mean age per title.

```{r}
data_imput <- data1 |> 
  #filter(source == "train") |> 
  mutate(title = str_extract(Name, ' ([A-Za-z]+)\\.'))

data_imput |> 
  group_by(title) |> 
  summarise(mean_age = mean(Age, na.rm = T),
            min_age = min(Age, na.rm = T),
            max_age = max(Age, na.rm = T),
            n = n()) |> 
  arrange(desc(n))
```

Some insights:

-   boys are called master, but unmarried (?) women are always called miss

    -   this is a bit of issue, because the mean imputation will assign too high age values for girls travelling with their parents.

    -   we could impute based on whether they travel alone or with their parents. This is likely only relevant for Miss, none of the other categories

-   some categories only have very few people –\> we should group them a bit more, pooling will probably give a more accurate mean imputation

```{r}
data_imput <- data_imput |> 
  mutate(title_group = case_when(
      title %in% c(" Capt.", " Col.", " Major.") ~ "Officer",
      title %in% c(" Jonkheer.", " Don.", " Sir.", " Dr.", " Rev.", " Countess.", " Lady.") ~ "Royalty",
      title %in% c(" Mme.", " Ms.", " Mrs.") ~ "Mrs",
      title %in% c(" Mlle.", " Miss.") ~ "Miss",
      title == " Mr." ~ "Mr",
      title == " Master." ~ "Master",
      TRUE ~ title
    ))
  
data_imput |> 
  group_by(title_group, Pclass) |> 
  summarise(mean_age = mean(Age, na.rm = T),
            n = n(), .groups = "drop") |> 
  arrange(title_group, Pclass) |> View()
```

```{r}
# add functions which do the mean imputation
# i would outsource these functions to RScripts to keep the document shorter

# Function to extract and group titles
create_title_groups <- function(data) {
  data |> 
    mutate(
      title = str_extract(Name, ' ([A-Za-z]+)\\.'),
      title_group = case_when(
        title %in% c(" Capt.", " Col.", " Major.") ~ "Officer",
        title %in% c(" Jonkheer.", " Don.", " Sir.", " Dr.", " Rev.", " Countess.", " Lady.") ~ "Royalty",
        title %in% c(" Mme.", " Ms.", " Mrs.") ~ "Mrs",
        title %in% c(" Mlle.", " Miss.") ~ "Miss",
        title == " Mr." ~ "Mr",
        title == " Master." ~ "Master",
        TRUE ~ title
      )
    )
}

# Function to calculate mean ages by title and class
calculate_group_means <- function(data) {
   data |> 
    filter(!is.na(Age)) |>
    group_by(title_group, Pclass) |>
    summarise(mean_age = mean(Age), .groups = "drop")
}

# Function to impute missing age
# Join on title group we create above and PClass
# IF Age is missing --> take imputed age (there is only one per title group and age)

impute_missing_ages <- function(data) {
  # Add title groups
  data_with_titles <- create_title_groups(data)
  
  # Calculate mean ages for each title group and passenger class
  group_means <- calculate_group_means(data_with_titles)
  
  # Impute missing ages
  data_with_titles |>
    left_join(group_means, by = c("title_group", "Pclass")) |>
    mutate(Age = coalesce(Age, mean_age)) |>
    select(-mean_age) # Remove the mean_age column after imputation
  
}
```

data1 \|\>

pull()

```{r}
#test imputation

data1 |> 
  mutate(title = str_extract(Name, ' ([A-Za-z]+)\\.')) |> 
  filter(str_detect(title, "Miss")) |> 
  group_by(Parch) |> 
  summarise(mean_age = mean(Age, na.rm = TRUE),
            n = n(),
            min_age = min(Age, na.rm = TRUE),
            max_age = max(Age, na.rm = TRUE))

data1 |> 
  mutate(title = str_extract(Name, ' ([A-Za-z]+)\\.')) |> 
  filter(str_detect(title, "Miss")) |> 
  group_by(Pclass, SibSp, Parch) |> 
  summarise(mean_age = mean(Age, na.rm = TRUE),
            n = n(),
            min_age = min(Age, na.rm = TRUE),
            max_age = max(Age, na.rm = TRUE))

data1 |> 
  mutate(title = str_extract(Name, ' ([A-Za-z]+)\\.')) |> 
  filter(str_detect(title, "Miss")) |> 
  ggplot(aes(x = Age)) +
  geom_histogram() +
  facet_wrap(~ Parch)

# this looks quite good. We should delete the imputation for miss based on Pclass and impute it according to Parch. 
```
